Wait-free Queues with Polygarithmic Step Complexity

### 2022-06-22 article/psuedocode.tex
fix true refresh

\Function{<Block, int, int>}{CreateBlock}{\sl{int} i} 
\Statex\cmt{Creates a block to be inserted into as $i$th \tt{block} in \tt{blocks}. Returns the created block as well as values read from each child's num\sub{propagated} field. These values are used for incrementing the children's num\sub{propagated} field if the block was appended to \tt{blocks} successfully.}
\State \tt{block newBlock= \Call{new}{\sl{block}}}
\State \tt{newBlock.group= num\sub{propagated}}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
\State \tt{index\sub{last}= dir.\head} \label{lastLine}
\State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
\State \tt{newBlock.end\sub{dir}= index\sub{last}}
\State \tt{block\sub{last}= dir.blocks[index\sub{last}]}
\State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
\State \cmt{\tt{newBlock} includes \tt{dir.blocks[index\sub{prev}+1..index\sub{last}]}.}
\State \tt{np\sub{dir}= dir.num\sub{propagated}}
\State \tt{newBlock.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}}
\State \tt{newBlock.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}}
\EndFor
%\State \tt{b.num\sub{enq}= b.num\sub{enq-left} + b.num\sub{enq-right}}
%\State \tt{b.num\sub{deq}= b.num\sub{deq-left} + b.num\sub{deq-right}}
%\State \tt{b.num= b.num\sub{enq} + b.num\sub{deq}}
%\State \tt{b.sum= n.blocks[i-1].sum + b.num}

@@

@@

\begin{definition}[Subblock]
Block \texttt{b} is a \emph{direct subblock} of \texttt{n.blocks[i]} if it is $\in$ \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}. Block \texttt{b} is a subblock of a \texttt{n.blocks[i]} if it is a direct subblock of it or subblock of a direct subblock of it. 
\end{definition}

@@

\begin{lemma}[trueRefresh] \label{lem::trueRefresh}
  Let $t_i$ be the time a \nf{n.Refresh()} operation is invoked and $t_t$ be the time it terminates. Suppose \nf{n.Refresh()}'s \nf{TryAppend(new, s)} returns \nf{true}, then \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{t}})}.
\end{lemma}
\begin{proof}
From Lemma \ref{lem::establishedOrder} we know that \nf{ops(EST\sub{n, t\sub{i}})}$\subseteq$\nf{ops(EST\sub{n, t\sub{t}})}. So it remains to show the operations of \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} - \nf{ops(EST\sub{n, t\sub{i}})}, which we call new operations, are all in \nf{ops(EST\sub{n, t\sub{t}})}. If \nf{TryAppend} returns true a block is appended to \nf{n}. From the code of the \nf{CreateBlock()} this block includes the established blocks in n's children at $t_i$. Since the head in \nf{CreateBlock()} is read after $t_i$. So the new operations are in the block appended to $n$.

%Suppose the previous true returning \nf{n.Refresh()} took place in time $(t_i^\prime-t_t^\prime)$. By induction on the number of succsuful \nf{Refresh()}'s \nf{n} we know that \nf{ops(EST\sub{n.left, t\sub{i}$^\prime$})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}$^\prime$})} $\in$ \nf{ops(EST\sub{n, t\sub{t}$^\prime$})}. Operations that are in \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} $\in$ \nf{ops(EST\sub{n, t\sub{t}})} but not in \nf{ops(EST\sub{n.left, t\sub{i}$^\prime$})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}$^\prime$})} $\in$ \nf{ops(EST\sub{n, t\sub{t}$^\prime$})} are in the block returned by \nf{n.CreateBlock()} by lemma 10. 
%By Lemma~\ref{lem::createBlock} \nf{new} contains \nf{n}'s childrens' established blocks after Line \ref{readHead} which is  appended to \nf{n.blocks[]} by \nf{TryAppend} in Line 313.
\end{proof}
\it{Mention this new block is establish in t\sub{t}. last sentece is not complete}

@@

\begin{lemma}[Get correctness] \label{get}
If \nf{n.blocks[b].num\sub{enq}$\geq$i} then \nf{n.GetENQ(b,i)} returns $E(n,b,i)$.
\end{lemma}
\begin{proof}
We are going to prove this lemma by induction on the height of the tree. The base case for the leaves of the tree is pretty straight forward. Since leaf blocks contain exactly one operation then only \nf{GetENQ(b,1)} can be called on leaves. \nf{leaf.GetENQ(b,1)} returns the operation stored in the $b$th block of leaf $l$. For non leaf nodes in Line 404 it is decided that the \nf{i}th enqueue in block \nf{b} of internal node \nf{n}resides in the left child or the right child of \nf{n}. By definition of $E(n,b)$ operations from the left child come  before the operations of the right child. Having \nf{sum\sub{enq}}, the prefix sum of the number of enqueues we can compute the direct subblock containing the enqueue we are finding for with binary search. Then \nf{n.child.GetENQ( block containing, order in the block)} is invoked which returns the correct operation by the hypothesis of the induction.
\end{proof}
\textit{I'm not sure it is going to be long and boring to talk about the parameters, since the reader can find out them.}

@@

%\begin{definition}
%An enqueue operation is \textit{finished} if its argument is returned by some process. A dequeue operation is \texttt{finished} if it returns \nf{null} or some value. Block \nf{b} is \textit{done} if all operations in \nf{ops(b)} are finished.
%\end{definition}
%\textit{Problem: we increment the \nf{num\sub{finished}} before returning and after the computing response. How to articulate the sentence above in a not confusing correct way?}
%
%\begin{lemma}[help]\label{help}
%After that \nf{TryAppend()} who is helping finishes, prefix for the blocks of \nf{root.blocks[root.FindMostRecentDone]} are done.
%\end{lemma}

\begin{lemma}[Computing SuperBlock]\label{superBlock}
After computing line \ref{computeSuper} of \nf{n.IndexDEQ(b,i)}, \nf{n.parent.blocks[superblock]} contains $D(n,b,i)$.
\end{lemma}
\begin{proof}
\begin{enumerate}
 \item Value read for \nf{super[b.group]} in line 418 is not null.\\$\blacktriangleright$
  Values \nf{c\textsubscript{dir}} read in lines 23, \nf{super} are set before incrementing in lines 26,27.

@@

\begin{lemma}[Search Ranges]\label{search}
  Preconditions of all invocation of \nf{BSearch} are satisfied.
\end{lemma}
\begin{proof}

Line 83: \nf{Get(i)} is called if the result of a dequeue is not null. The search is among all blocks in the root.

Line 88: This search tries to find the ith enqueue, knowing that it is in the left child. Search is done over the left subblocks. The start and end of the range are followed by definition. Line 92 is the same.

Line 101: Here, the goal is to find the superblock. We know the distance between answer and the \nf{super[i]} is at most $p$, since at most $p$ processes could die.

\end{proof}

@@

\begin{lemma}[Root search range]\label{rootRange}
  \nf{root.size-root.FindMostRecentDone()} is $O(p^2+q)$, which $p$ is \# processes and $q$ is the \size of the queue.
\end{lemma}

@@

\end{document}

>>>>

\Function{<Block, int, int>}{CreateBlock}{\sl{int} i} 
\Statex\cmt{Creates a block to be inserted into as $i$th \tt{block} in \tt{blocks}. Returns the created block as well as values read from each child's num\sub{propagated} field. These values are used for incrementing the children's num\sub{propagated} field if the block was appended to \tt{blocks} successfully.}
\State \tt{block newBlock= \Call{new}{\sl{block}}}
\State \tt{newBlock.group= num\sub{propagated}}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
\State \tt{index\sub{last}= dir.\head} \label{lastLine}
\State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
\State \tt{newBlock.end\sub{dir}= index\sub{last}} \label{endDefLine}
\State \tt{block\sub{last}= dir.blocks[index\sub{last}]}
\State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
\State \cmt{\tt{newBlock} includes \tt{dir.blocks[index\sub{prev}+1..index\sub{last}]}.}
\State \tt{np\sub{dir}= dir.num\sub{propagated}}
\State \tt{newBlock.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}}
\State \tt{newBlock.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}}
\EndFor
%\State \tt{b.num\sub{enq}= b.num\sub{enq-left} + b.num\sub{enq-right}}
%\State \tt{b.num\sub{deq}= b.num\sub{deq-left} + b.num\sub{deq-right}}
%\State \tt{b.num= b.num\sub{enq} + b.num\sub{deq}}
%\State \tt{b.sum= n.blocks[i-1].sum + b.num}

@@

\framebox[1.1\width]{Question} How to remove the notion of time? To say pre(n,i) contains n.blocks[0..i] instead of EST(n,t) which head=i at time t. Is it good?

@@

\begin{definition}[Subblock]
Block \texttt{b} is a \emph{direct subblock} of \texttt{n.blocks[i]} if it is $\in$ \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]} (See line \ref{endDefLine} for the defined range). Block \texttt{b} is a subblock of a \texttt{n.blocks[i]} if it is a direct subblock of it or subblock of a direct subblock of it. 
\end{definition}

@@

\begin{lemma}[head Increment] \label{lem::headInc}
If an n.Refresh  instance reaches Line \ref{cas} instance and reads head=h (\ref{readHead}) after it terminates head is greater than h.
\end{lemma}
\begin{proof}
If Line \ref{incrementHead1} or \ref{incrementHead1} succeeded the claim holds, otherwise another process has incremented the head.
\end{proof}

\begin{lemma}[trueRefresh] \label{lem::trueRefresh}
  Let $t_i$ be the time an instance of \nf{n.Refresh()} is invoked and $t_t$ be the time it terminates. Suppose the \nf{TryAppend(new, s)} of the \nf{n.Refresh()}  returns \nf{true}, then \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{t}})}.
\end{lemma}
\begin{proof}
From Lemma \ref{lem::establishedOrder} we know that \nf{ops(EST\sub{n, t\sub{i}})}$\subseteq$\nf{ops(EST\sub{n, t\sub{t}})}.
So it remains to show the operations of \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} - \nf{ops(EST\sub{n, t\sub{i}})}, which we call \it{new operations}, are all in \nf{ops(EST\sub{n, t\sub{t}})}. 
If \nf{TryAppend} returns true a block \nf{new} is written into \nf{n.blocks[h]} (Line \ref{cas}).
We show \nf{ops(EST\sub{n.left, t\sub{i}})}$\subseteq$\nf{ops(EST\sub{n, t\sub{t}})}. The proof for the right child's claim is the same. Let \nf{n.left.head} at \nf{t\sub{i}} be \nf{hli}. Let \nf{n.Refresh()} read \nf{head} equal to $h$(Line \ref{readHead}). By the lines \ref{prevLine},\ref{lastLine} the \nf{new} block in \nf{n.blocks[h]} contains \nf{n.left.blocks[ n.blocks[h-1].end\sub{left}+1..left.head]}. Since \nf{left.head} is read after \nf{t\sub{i}} then \nf{ops(EST\sub{n.left, t\sub{i}})}$\subseteq$\nf{ops(n.left.blocks [0..left.head])}. By Lemma \ref{lem::establishedOrder} \nf{ops(n.left.blocks[0..n.blocks[h-1].end\sub{left}])}$\subseteq$\nf{ops(EST\sub{n, t\sub{i}})}$\subseteq$\nf{ops(EST\sub{n, t\sub{t}})}. Since after line \ref{incrementHead2} we are sure that the \nf{head} is incremented (Lemma\ref{lem::headInc}) and \nf{n.head=h+1} at $t_t$ so the new block is established at $t_t$ and the new block contains the new operations which is what we wanted to show.
% From the code of the \nf{CreateBlock()} this block includes the established blocks in n's children at $t_i$. Since the head in \nf{CreateBlock()} is read after $t_i$. So the new operations are in the block appended to $n$.

%Suppose the previous true returning \nf{n.Refresh()} took place in time $(t_i^\prime-t_t^\prime)$. By induction on the number of succsuful \nf{Refresh()}'s \nf{n} we know that \nf{ops(EST\sub{n.left, t\sub{i}$^\prime$})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}$^\prime$})} $\in$ \nf{ops(EST\sub{n, t\sub{t}$^\prime$})}. Operations that are in \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} $\in$ \nf{ops(EST\sub{n, t\sub{t}})} but not in \nf{ops(EST\sub{n.left, t\sub{i}$^\prime$})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}$^\prime$})} $\in$ \nf{ops(EST\sub{n, t\sub{t}$^\prime$})} are in the block returned by \nf{n.CreateBlock()} by lemma 10. 
%By Lemma~\ref{lem::createBlock} \nf{new} contains \nf{n}'s childrens' established blocks after Line \ref{readHead} which is  appended to \nf{n.blocks[]} by \nf{TryAppend} in Line 313.
\end{proof}
\begin{figure}[hbt]
  \center\includegraphics[width=5in]{pics/trueRefresh}
  \caption{New established operations of the left child are in the new block.}
\end{figure}
%\it{Mention this new block is establish in t\sub{t}. last sentece is not complete}

@@

Preconditions of all invocation of \nf{BSearch} are satisfied.
\begin{lemma}[Get correctness] \label{get}
If \nf{n.blocks[b].num\sub{enq}$\geq$i} then \nf{n.GetENQ(b,i)} returns $E(n,b,i)$.
\end{lemma}
\begin{proof}
We are going to prove this lemma by induction on the height of the tree. The base case for the leaves of the tree is pretty straight forward. Since leaf blocks contain exactly one operation then only \nf{GetENQ(b,1)} can be called on leaves. \nf{leaf.GetENQ(b,1)} returns the operation stored in the $b$th block of leaf $l$. For non leaf nodes in Line 404 it is decided that the \nf{i}th enqueue in block \nf{b} of internal node \nf{n}resides in the left child or the right child of \nf{n}. By definition of $E(n,b)$ operations from the left child come  before the operations of the right child. Having \nf{sum\sub{enq}}, the prefix sum of the number of enqueues we can compute the direct subblock containing the enqueue we are finding for with binary search. Then \nf{n.child.GetENQ( block containing, order in the block)} is invoked which returns the correct operation by the hypothesis of the induction.
\end{proof}
\textit{I'm not sure it is going to be long and boring to talk about the parameters, since the reader can find out them.}

@@

%\begin{definition}
%An enqueue operation is \textit{finished} if its argument is returned by some process. A dequeue operation is \texttt{finished} if it returns \nf{null} or some value. Block \nf{b} is \textit{done} if all operations in \nf{ops(b)} are finished.
%\end{definition}
%\textit{Problem: we increment the \nf{num\sub{finished}} before returning and after the computing response. How to articulate the sentence above in a not confusing correct way?}
%
%\begin{lemma}[help]\label{help}
%After that \nf{TryAppend()} who is helping finishes, prefix for the blocks of \nf{root.blocks[root.FindMostRecentDone]} are done.
%\end{lemma}
Preconditions of all invocation of \nf{BSearch} are satisfied.
\begin{lemma}[Computing SuperBlock]\label{superBlock}
After computing line \ref{computeSuper} of \nf{n.IndexDEQ(b,i)}, \nf{n.parent.blocks[superblock]} contains $D(n,b,i)$.
\end{lemma}
\begin{proof}
\begin{enumerate}
 \item Value read for \nf{super[b.group]} in line 418 is not null.\\$\blacktriangleright$
  Values \nf{c\textsubscript{dir}} read in lines 23, \nf{super} are set before incrementing in lines 26,27.

@@

%\begin{lemma}[Search Ranges]\label{search}
%  Preconditions of all invocation of \nf{BSearch} are satisfied.
%\end{lemma}
%\begin{proof}
%  
%Line 83: \nf{Get(i)} is called if the result of a dequeue is not null. The search is among all blocks in the root.
%
%Line 88: This search tries to find the ith enqueue, knowing that it is in the left child. Search is done over the left subblocks. The start and end of the range are followed by definition. Line 92 is the same.
%
%Line 101: Here, the goal is to find the superblock. We know the distance between answer and the \nf{super[i]} is at most $p$, since at most $p$ processes could die.
%
%\end{proof}

@@

@@

\begin{lemma}[Root search range]\label{rootRange}
  \nf{root.size-root.FindMostRecentDone()} is $O(p^2+q)$, which $p$ is \# processes and $q$ is the \size of the queue.
\end{lemma}
\end{document}
